---
title: "04_Biodiversity"
author: "Jayden William Kgobe"
date: "`r Sys.Date()`"
output:output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../../figures/04_Biodiversity/",
                      fig.align = "center")
```


# Goals 

Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the Hill numbers as it relates to stations and salinity (in PSU).
4. Test and then answer the scientific question at hand! (Remember, we're doing biology!)

## Inputs 

1. We will need the `preprocessed_physeq` that we created in `analysis/03_PreProcessing.Rmd`, which has the chloroplasts, mitochondria, and any controls removed. 

## Outputs 

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to include in our scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests conveying the measured and quantified changes and patterns in biodiversity.

# Scientific Question
How does rumen microbial alpha diversity vary across samples in this dataset, and to what extent is this variation influenced by sequencing depth

## Question & Hypotheses

### Null Hypothesis
There is no meaningful variation in rumen microbial alpha diversity across samples beyond random variation, and alpha diversity is not associated with sequencing depth.

### Alternative Hypothesis
Rumen microbial alpha diversity varies across samples, and this variation is not solely explained by sequencing depth.


# Set up the Environment 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
set.seed(2356728)
```
## Load Packages & Colors 

```{r load-packages-colors}
# load packages
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

# Load in colors for plotting 
source("code/colors_and_shapes.R")
```

## Load in Data 

```{r load-data}
load("data/03_PreProcessing/preprocessed_phyloseq.RData")

# take a look at it! 
preprocessed_physeq

# Intuition check on seq depth
min(sample_sums(preprocessed_physeq))

# Make a metadata dataframe 
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()

# view of it
glimpse(metadata_df)
```
# Biodiversity Hill Numbers with iNEXT


## Hill Number Equation

Next, we will calculate the biodiversity with the [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) package. iNEXT focuses on three measures of biodiversity via the Hill numbers. Hill numbers have the order q: 

$$
{}^qD = \left( \sum_{i=1}^{S} p_i^q \right)^{1/(1 - q)}, \quad q \geq 0, \quad q \neq 1.
$$

But when *q = 1*, then the equation is: 

$$
{}^1D = \lim_{q \to 1} {}^qD = \exp\left( - \sum_{i=1}^{S} p_i \log p_i \right) = \exp(H_{sh}).
$$


**Where:**

- \( {}^qD \) is the *Hill number* or *diversity of order \( q \)*, a measure of species diversity.
- \( q \) is the *order of diversity*, which determines how sensitive the measure is to species relative abundances:
  - \( q = 0 \): all species are weighted equally or **species richness**,
  - \( q = 1 \): corresponds to the exponential of **Shannon entropy**,
  - \( q = 2 \): corresponds to the **inverse of Simpson’s index**.
- \( S \) is the *total number of species* in the community.
- \( p_i \) is the *relative abundance* of species \( i \), i.e., the proportion of individuals in the community belonging to species \( i \).

## Microbial Example 🦠

Imagine you're studying the microbial diversity in three sourdough starters from three different bakers: A, B, and C. Each one contains a mix of **ASVs**. *But how diverse are they, really?* Here,'s a great application of the **Hill numbers** to find out, varying the value of \( q \).

### When \( q = 0 \): Counting *All* ASVs Equally  
> *Species Richness* → \( {}^0D \)

Here, we ignore how common or rare each ASV is. Every ASV counts equally — even ones that appear just once. For example, this is like walking into a library and simply counting how many unique book titles are on the shelves—not how many copies of each exist, just the number of different titles. So, even if there’s only one copy of most books and 100 copies of one, they all count the same. 

Here, in the context of microbial communities, we are pondering: **“How many different ASVs are present?”** We do not weight them by abundance/evenness in any way. Rather, we focus on a binary measures of 0s and 1s that represent presence (1) and absence (0).

- Baker A’s starter has 10 unique ASVs: \( {}^0D_{\text{A}} = 10 \)    
- Baker B’s has 7 unique ASVs: \( {}^0D_{\text{B}} = 7 \)  
- Baker C’s has 15 unique ASVs: \( {}^0D_{\text{C}} = 15 \)  


### When \( q = 1 \): Balancing Rarity and Commonness  
> *Shannon Diversity* → \( {}^1D \)

Now we ask: **"How evenly are the ASVs distributed?"** Rare ASVs matter, but not as much as common ones. If we revisit the library analogy, now imagine you’re randomly picking books off the shelves to read. Some books have many more copies, so you’re more likely to pick those — but you still have a chance to encounter the rare ones. This measure reflects both richness and evenness — it tells you the effective number of “common” book titles based on their relative frequency. `q = 1` gives you a sense of how diverse your reading experience would be, considering popular vs. rare titles.

- Baker A’s 10 ASVs are 90% dominated by just one ASV: \( {}^1D_{\text{A}} \approx 2 \)  
- Baker B’s 7 ASVs are fairly even: \( {}^1D_{\text{B}} \approx 6.8 \)  
- Baker C’s 15 ASVs include 3 very dominant ones: \( {}^1D_{\text{C}} \approx 5 \)  

Even though Baker C and A have more ASVs in terms of richness in \( {}^0D \), **Baker B** has the highest \( {}^1D \) due to better balance between the abundances of its seven ASVs. 

### When \( q = 2 \): Focus on the Most Abundant ASVs  
> *Simpson-like Diversity* → \( {}^2D \)

Now we modify our question to wonder: **How many ASVs dominate each sample?** Here, rare ASVs hardly matter. For example, this is like walking into the library and only noticing the most abundant book titles, because they take up entire shelves. The rare titles are barely visible. Therefore, this measure emphasizes dominant species—those with many copies. `q = 2` tells us the effective number of dominant book titles, downplaying the presence of rare ones.

- Baker A’s starter is almost all one ASV of 10 total ASVs: \( {}^2D_{\text{A}} \approx 1.2 \)  
- Baker B’s 7 ASVs are balanced: \( {}^2D_{\text{B}} \approx 6.5 \)   
- Baker C’s has 3 dominant ASVs out of 15: \( {}^2D_{\text{C}} \approx 3.5 \)  

This tells us: **Baker B** has the most diverse community in terms of its *dominant* community members.


---

## Hill Number Summary Table

| Value of \( q \) | Emphasis                    | Hill Number \( {}^qD \) | Interpretation                              |
|------------------|-----------------------------|-------------------------|----------------------------------------------|
| \( q = 0 \)      | Every ASV counts equally     | \( {}^0D \)             | ASV richness — a simple count                |
| \( q = 1 \)      | Balance of common & rare     | \( {}^1D \)             | Shannon Diversity — balanced view            |
| \( q = 2 \)      | Dominant ASVs matter most    | \( {}^2D \)             | Simpson Diversity — focus on major players        |

---

So, hopefully you can see that using Hill numbers like \( {}^qD \) gives you a powerful, flexible way to describe microbial diversity — from just counting ASVs to understanding community structure. Next, we will calculate the biodiversity with the [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) package. iNEXT focuses on three measures of biodiversity via the Hill numbers. For each Hill NUmber [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) uses the observed abundance of species within a sample to compute diversity estimates and the associated 95% confidence intervals for the following two types of rarefaction and extrapolation (R/E):

1. **Sample-size-based (or size-based) R/E curves**: iNEXT computes diversity estimates for
rarefied and extrapolated samples up to an appropriate size (2x the total number of individuals/sequences in that sample). This type of sampling curve plots the diversity estimates with respect to sample size.
2. **Coverage‐based R/E curves:** iNEXT computes diversity estimates for rarefied and extrapolated samples based on a standardized level of sample completeness (as measured by sample coverage) up to an appropriate coverage value. This type of sampling curve plots the diversity estimates with respect to sample coverage. **NOTE: Coverage is inapplicable to DADA2 datasets, as they lack singletons** (this is also why we can’t use Amy Willis’s breakaway approach). 

*An important note:* We’ll be using the size_based output. Our lab has previously been suspicious of the Asymptotic Estimator. Therefore we will use the size_based approach here, with the caveat that we will very rigorously confirm that we’ve approached saturation for all of our samples! 


> Warning! iNEXT takes a while to run, especially if your samples are deeply sequenced or if you have many samples! It is wise to calculate iNEXT early and then come back to the rest of the home work later on.

```{r iNEXT-biodiversity}
# prepare the input data 
# Species in rows and the samples in columns 
iNEXT_input_df <- 
  preprocessed_physeq %>%
  otu_table() %>%
  data.frame()

# Inpect 
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

# Run iNEXT 
# Remember to set the seed!
# ASVs in ROWs, samples in COLUMNS 
#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                  q = c(0, 1, 2),
#                   datatype = "abundance")

# Save the object for later! 
# Make the directory first in shell! Otherwise there will be an error
#save(iNEXT_data, file = "data/04_Biodiversity/iNEXT_data.RData")
```

Since iNEXT takes a while to run, we will load in the data that was caluculated before! 

```{r load-iNEXT-data}
load("data/04_Biodiversity/iNEXT_data.RData")

# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)
```


# Rarefaction Curves 

## `ggiNEXT` Rarefaction Curves 

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()

# Check
head(color_df)

# Rename the column 
colnames(color_df)[1] <- "names"

# Check
head(color_df)
metadata_df

colnames(color_df)

colnames(metadata_df)

# Make a helper dataframe for plotting with colors 
group_colors <- c("China" = "red4")

iNEXT_color_df <- 
  color_df %>%
  mutate(
    names = gsub(pattern = "[.]", replace = "-", x = names),
    names = gsub(pattern = "^X", replace = "", x = names)
  ) %>%
  rename(SampleID = names) %>%
  left_join(metadata_df, by = "SampleID") %>%
  left_join(
    data.frame(
      group_colors = group_colors,
      geo_loc_name_country = names(group_colors)
    ),
    by = "geo_loc_name_country"
  )

head(iNEXT_color_df)
```

```{r plot-iNEXT-rarefaction, fig.width=6, fig.height=3.5}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$group_colors) + 
  scale_fill_manual(values = iNEXT_color_df$group_colors) + 
  theme_bw() + 
  theme(legend.position = "none") + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") 
```
Remembering that an Order q of:  

  - 0 = Richness/ Number of Total taxa
  - 1 = Exponential Shannon / Number of "Common" taxa
  - 2 = Inverse Simpson / Number of "Dominant" taxa 
  
*Also note that iNEXT will extrapolate the number of sequences to double the sequencing depth of the sample!* This extrapolation is represented by the dotted line, which extends to 2 x the library size of the sample. 

Let's draw some conclusions from the plot above:  

1. The best gift of the Hill Numbers: *All three diversity measures are on the same x-axis: The Effective Number of Species or ASVs.*  
2. As q increases from 0 to 1 to 2, the abundance of the species is weighted more, so the rarefaction curve plateaus earlier. This is important because it **indicates that richness is the most sensitive to sequencing depth!** However, Shannon and even less sensitive is Simpson.  
3. Most of the sample rarefaction curves appear to plateau but we need to look deeper to truly confirm this.  
4. There appears to be a sample with much lower richness than the rest of the dataset. Let's keep our eyes on this sample...  

Ok, this plot is nice, but since it's autogenerated with the `ggiNEXT` function, I'd like some more freedom to plot and explore the data further. Let's manually plot the data that we pull out from the iNEXT data.

## Manual Rarefaction Curves 

```{r manual-rarefaction, fig.width=6, fig.height=5}
str(iNEXT_data)

iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # fix the samples to merge with metadata 
  mutate(
    names = gsub("[.]", "-", names),
    names = gsub("X", "", names)
  ) %>%
  # join with the metadata
  left_join(metadata_df, by = c("names" = "SampleID"))

# Inspection
dim(iNEXT_manual_df)

# Manually plot rarefaction curves 
iNEXT_manual_df %>%
  # Filter out rows that are calculated with actual sequences (no extrapolated)
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot! 
  ggplot(aes(x = m, y = qD, color = geo_loc_name_country, group = names)) + 
  geom_line() + 
  # Facet by station 
  facet_grid(Order.q~geo_loc_name_country, scales = "free") + 
  scale_color_manual(values = salinity_colors) + 
  theme_bw() + 
  labs(y = "Effective Number of ASVs",
       x = "Number of Sequences") + 
  theme(legend.position = "bottom")
```

<span style="color: red;">INTERPRETATION #1:  What can you conclude from the ggiNEXT and manual rarefaction curve plots? Are there "enough" sequences to analyze the samples? Is it valid to make any conclusions about species richness when q = 0? What about when q = 1? Or q = 2? Why or why not?</span>

**Summary of Observations**
Both the ggiNEXT and manual rarefaction curves, most samples show a rapid initial increase followed by a clear plateau, particularly at lower sequencing depths. This indicates that:

- The majority of microbial diversity is captured early and any dditional sequencing yields diminishing returns
- Therefore, sequencing depth is generally sufficient for downstream analysis

**Observation of Richness: (q = 0)**
- Some of the curves at q = 0 do not fully plateau (or maybe I am being too pedantic); they seem to to continue to rise at higher sequence depths
- This observation would imply that richness is highly sensitive to rare taxa therefore, ess reliable to make definitive conclusions about species richness at q = 0, because richness is likely underestimated.

**Observation of Shannon Diversity: (q = 1)**
- The curves in this case seem to plateau more constistnetly actoss samples.
- It is valid to make conclusions about diversity at q = 1, as sequencing depth appears sufficient to capture the majority of ecologically relevant diversity.

**Observation of Simpson Diversity: (q = 2)**
- Much more steeper curves, so the plateaus in the curves appear much more quickly. Since Simpson diversity emphasizes dominant taxa, rare taxa have little influence in this community
- It is valid to make conclusions at q = 2, since dominant community structure is well captured even at lower sequencing depths.

**Conclusions**
- Sequencing depth is adequate...yay!!!
- However, Confidence in diversity estimates increases with higher q values:
    * q = 0: least reliable; since it is sensitive to rare taxa
    * q = 1: reliable
    * q = 2: reliable

# Statististical Testing 

So, how do we actually quantify the changes in biodiversity that we may observe between the samples? That takes us into statistics! 


## Parametric vs. Non-Parametric Statistics??

Understanding the difference between **parametric** and **non-parametric** statistics is essential for choosing the right test for your data, especially in microbial ecology where data often violate assumptions of normality.

*So, what's the difference between the two?*

| Feature                     | Parametric                                | Non-Parametric                          |
|----------------------------|--------------------------------------------|-----------------------------------------|
| **Assumptions**            | Assumes a known distribution (usually normal) | Fewer assumptions about distribution    |
| **Data Type**              | Continuous (interval/ratio)                | Ordinal or non-normal continuous        |
| **Examples**               | t-test, ANOVA, linear regression           | Wilcoxon, Kruskal-Wallis, Spearman      |
| **Estimates Based On**     | Parameters like mean and variance          | Ranks or medians                        |
| **Sensitive to Outliers?**| Yes                                        | Less sensitive                          |
| **Statistical Power**      | Higher if assumptions are met              | Lower but more robust to violations     |
### Parametric Statistics

Parametric tests assume that the data follow a specific distribution (usually a normal/Gaussian distribution). They are more statistically powerful **if those assumptions are met**. Some examples include, comparing means with a t-test or ANOVA or modeling relationships using linear regression. 

```
# Example t-test on alpha diversity between two treatment groups (Shannon index)
t.test(shannon ~ treatment, data = my_data)
```

### Non-Parametric Statistics

Non-parametric tests do not assume a specific data distribution and are more robust to outliers. However, they are less robust compared to parametric statistics because of this flexibility. Some examples include wilcoxon tests to compare diversity between sample types. Non-parametric tests are ideal when:

- Data are skewed or non-normal
- Sample sizes are small
- Outliers are present
- You’re analyzing ranks (*e.g.,* ordinal data) or medians

```
# Example Wilcoxon test on alpha diversity between two treatment groups (Shannon index)
wilcox.test(shannon ~ treatment, data = my_data)
```

## How do I test if my data is "normal"? 

Typically people perform a **Shapiro-Wilk Test** on their data. The Shapiro-Wilk test is a statistical test used to check whether a sample of data comes from a normally distributed population.

The Shapiro-Wilk test calculates a **W statistic** by comparing the order statistics (sorted values) of your data with those expected from a normal distribution.

- **W ≈ 1** → data is close to normal.
- **W < 1** → increasing deviation from normality.

What hypothses does the Shapiro-Wilk Test? 

- *Null hypothesis (H₀):* The data are normally distributed.
    - *p-value > 0.05*: Fail to reject H₀ and the data appears normal.
- *Alternative hypothesis (H₁):* The data are *not* normally distributed.
    - *p-value < 0.05*: We can reject the H₀, which will indicate that the data is **not normal**.


**Assumption would be to assume not normally distributed because the Shapiro-Wilk test is very sensitive to small datasets like mine**

# Continuous Relationships 

Here, we will test the continuous variable 

## Spearman vs. Linear Model: What’s the Difference?

| Feature                  | **Spearman Correlation**                           | **Linear Model (`lm()`)**                               |
|--------------------------|----------------------------------------------------|----------------------------------------------------------|
| **Type of Relationship** | Monotonic (increasing or decreasing)              | Linear (straight-line relationship)                      |
| **Assumptions**          | Non-parametric (no distribution assumptions)      | Parametric (normality, linearity, homoscedasticity)      |
| **Input Data**           | Ordinal, ranked, or continuous                    | Continuous (can include categorical predictors too)      |
| **What it Measures**     | Strength and direction of monotonic relationship  | Effect size, direction, significance of predictors        |
| **Output**               | Correlation coefficient (ρ) and p-value           | Intercept, slope(s), p-values, R², residuals, etc.       |
| **Resistant to Outliers?** | More robust                                     | Sensitive to outliers                                    |
| **Function in R**        | `cor.test(x, y, method = "spearman")`             | `lm(y ~ x)`                                              |
# Continuous Variables 

```{r comptute-alpha}



#load otu_table
load("data/03_PreProcessing/preprocessed_phyloseq.RData")

noChlorosMitos_physeq <- get("preprocessed_physeq")

#Intuition Check
class(noChlorosMitos_physeq)
noChlorosMitos_physeq
phyloseq::otu_table(noChlorosMitos_physeq)

# 2. Alpha diversity
alpha_df <- 
  phyloseq::estimate_richness(
    noChlorosMitos_physeq,
    measures = c("Observed", "Shannon", "Simpson")
  ) %>%
  tibble::rownames_to_column(var = "SampleID")

# 3. Sequencing depth
seqSums_df <- 
  data.frame(TotalSeqs = phyloseq::sample_sums(noChlorosMitos_physeq)) %>%
  tibble::rownames_to_column(var = "SampleID")

# 4. Combine
obs_div_df <- 
  alpha_df %>%
  dplyr::left_join(seqSums_df, by = "SampleID") %>%
  tidyr::pivot_longer(
    cols = c("Observed", "Shannon", "Simpson"),
    names_to = "Order.q",
    values_to = "qD"
  )

# 5. Label q values
obs_div_df$Order.q <- dplyr::recode(
  obs_div_df$Order.q,
  "Observed" = "q = 0 (Richness)",
  "Shannon"  = "q = 1 (Shannon)",
  "Simpson"  = "q = 2 (Simpson)"
)

glimpse(obs_div_df)
    

```

# Environmental Data

```{r env-data}
# Create dataframe of continuous variables 
env_df <- 
  obs_div_df %>%
  dplyr::select(SampleID, TotalSeqs, qD)

# Pair-wise interactions between variables
pairs(dplyr::select(env_df, -SampleID), upper.panel = NULL)
```


**NOTE!!**
- There seems to be two groups in sample:
    * Ones with a positive correlation
    * One with no correlation
    
# Continuous Variable Plot

```{r plot-cont-var}
# How does biodiversity change over the sequencing depth gradient?
obs_div_df %>%
  ggplot(aes(x = TotalSeqs, y = qD)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_point() + 
  labs(x = "Sequencing Depth (Total Reads)", y = "Alpha Diversity") +
  stat_smooth(method = "lm", formula = y ~ x) + 
  theme_bw()
```

### Explore Richness More

```{r richness-vs-salinity}
# Make a richness data frame
richness_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == "q = 0 (Richness)")

# Actually run the linear model
rich_vs_depth_lm <- 
  lm(qD ~ TotalSeqs, data = richness_df)

# Show me the model results
rich_vs_depth_lm

# Type?
class(rich_vs_depth_lm)

# Summary of lm
summary(rich_vs_depth_lm)
```
In the output:

- **Coefficients** refer to the 𝛽’s
- **Estimate** is the estimate of each coefficient
- **Std. Error** is the standard error of the estimate
- **t value** is the coefficient divided by its standard error
- **Pr(>|t|)** is the p-value for the coefficient
- The **residual standard error** is the estimate of the variance of 𝜖
- **Degrees of freedom** is the sample size minus # of coefficients estimated
- **R-squared** is (roughly) the proportion of variance in the outcome explained by the model, instead we should always report the **adjusted R-squared**, which takes into penalizes the addition of unnecessary predictors, which provides a more accurate measure of model fit.
    - Note that the R-squared value will always increase/stay the same when you add more predictors, even if they are not relevant!
- The **F-statistic** compares the fit of the model as a whole to the null model (with no covariates)
    - When comparing models, we always want to take the model with the higher F-statistic because it indicates a stronger relationship between the predictor varaibles and the dependent variables, suggesting that the model is better. 
    
```{r plot-richness-sal, fig.height=4, fig.width=4, warning = FALSE}

richness_df %>% 
  ggplot(aes(x = TotalSeqs, y = qD)) + 
  geom_point() + 
  stat_smooth(method = "lm") + 
  labs(x = "Sequencing Depth (Total Reads)", y = "# of ASVs") + 
  theme_bw() + 
  geom_label(aes(x = max(TotalSeqs)*0.6, y = max(qD)*0.75), hjust = 0, 
             label = paste("Adj R2 = ",signif(summary(rich_vs_depth_lm)$adj.r.squared, 2),
                           "\nIntercept =",signif(rich_vs_depth_lm$coef[[1]],3),
                           " \nSlope =",signif(rich_vs_depth_lm$coef[[2]], 2),
                           " \nP =",signif(summary(rich_vs_depth_lm)$coef[2,4], 2)))
```
## Spearman Correlation
```{r rich-vs-salinity-spearman, fig.width=4, fig.height=4}
richness_df %>% 
  ggplot(aes(x = TotalSeqs, y = qD)) + 
  geom_point() + 
  labs(x = "Sequencing Depth (Total Reads)", y = "# of ASVs") + 
  theme_bw() + 
  ggpubr::stat_cor(method = "spearman")
```

<span style="color: red;">INTERPRETATION #4:  If you have continuous data, what can you conclude regarding the role that all ASVs (richess, q = 0) versus common ASVs (Shannon, q = 1) versus dominant ASVs (Simpson, q = 2) have on the biodiversity of your samples as it relates to your variable of interest?</span>

**1. Richness: q = 0**
- There is a strong positive linear relationship between sequencing depth and richness. This is because: 
  *Slope = 0.026: richness increases as sequencing depth increases
  *Adjusted R² = 0.69: sequencing depth explains a large portion of variation
  *p < 2.2e-16: highly significant (I am doubtful of this p value...is it trully that low?)
- This implies richness is highly sensitive to sequencing depth. As more reads are collected, more rare ASVs are detected. All ASVs (including rare taxa) strongly influence biodiversity estimates, but this relationship is largely driven by sequencing effort rather than true biological differences.

**Shannon: q = 1**
- There is a weaker, but, positive relationship as seen by a less steep slope. This implies common ASVs contribute to biodiversity patterns, but are less dependent on sequencing depth compared to richness.

**Simpson: q = 2**
- he relationship between sequencing depth and diversity is very weak as seen by flat lines. This emphasizes dominant taxa which are captured even at lower sequencing depths
- Dominant ASVs have little dependence on sequencing depth, indicating that the core community structure is consistently captured across samples.



<span style="color: red;">INTERPRETATION #5:  How did you statistically test your scientific question in Interpretation #4? Why?</span>

**Linear Model**
- We quantified the direction and strength of the relationship in order to estimate the effect size (slope) as wellas determine whether the relationship was statistically significant
- here is a strong, statistically significant positive relationship between sequencing depth and observed richness. Observed by 
  *Slope = 0.0256: richness increases with sequencing depth
  *Adjusted R² = 0.69: sequencing depth explains ~69% of variation in richness

**Spearman Correlation Test**
- There is a strong monotonic relationship between sequencing depth and richness (ρ ≈ 0.87).

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```

