---
title: "04_Biodiversity"
author: "Jayden William Kgobe"
date: "`r Sys.Date()`"
output:output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../../figures/04_Biodiversity/",
                      fig.align = "center")
```


# Goals 

Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the Hill numbers as it relates to stations and salinity (in PSU).
4. Test and then answer the scientific question at hand! (Remember, we're doing biology!)

## Inputs 

1. We will need the `preprocessed_physeq` that we created in `analysis/03_PreProcessing.Rmd`, which has the chloroplasts, mitochondria, and any controls removed. 

## Outputs 

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to include in our scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests conveying the measured and quantified changes and patterns in biodiversity.

# Scientific Question

## Question & Hypotheses


# Set up the Environment 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
set.seed(2356728)
```
## Load Packages & Colors 

```{r load-packages-colors}
# load packages
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

# Load in colors for plotting 
source("code/colors_and_shapes.R")
```

## Load in Data 

```{r load-data}
load("data/03_PreProcessing/preprocessed_phyloseq.RData")

# take a look at it! 
preprocessed_physeq

# Intuition check on seq depth
min(sample_sums(preprocessed_physeq))

# Make a metadata dataframe 
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()

# view of it
glimpse(metadata_df)
```
# Biodiversity Hill Numbers with iNEXT


## Hill Number Equation

Next, we will calculate the biodiversity with the [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) package. iNEXT focuses on three measures of biodiversity via the Hill numbers. Hill numbers have the order q: 

$$
{}^qD = \left( \sum_{i=1}^{S} p_i^q \right)^{1/(1 - q)}, \quad q \geq 0, \quad q \neq 1.
$$

But when *q = 1*, then the equation is: 

$$
{}^1D = \lim_{q \to 1} {}^qD = \exp\left( - \sum_{i=1}^{S} p_i \log p_i \right) = \exp(H_{sh}).
$$


**Where:**

- \( {}^qD \) is the *Hill number* or *diversity of order \( q \)*, a measure of species diversity.
- \( q \) is the *order of diversity*, which determines how sensitive the measure is to species relative abundances:
  - \( q = 0 \): all species are weighted equally or **species richness**,
  - \( q = 1 \): corresponds to the exponential of **Shannon entropy**,
  - \( q = 2 \): corresponds to the **inverse of Simpson’s index**.
- \( S \) is the *total number of species* in the community.
- \( p_i \) is the *relative abundance* of species \( i \), i.e., the proportion of individuals in the community belonging to species \( i \).

## Microbial Example 🦠

Imagine you're studying the microbial diversity in three sourdough starters from three different bakers: A, B, and C. Each one contains a mix of **ASVs**. *But how diverse are they, really?* Here,'s a great application of the **Hill numbers** to find out, varying the value of \( q \).

### When \( q = 0 \): Counting *All* ASVs Equally  
> *Species Richness* → \( {}^0D \)

Here, we ignore how common or rare each ASV is. Every ASV counts equally — even ones that appear just once. For example, this is like walking into a library and simply counting how many unique book titles are on the shelves—not how many copies of each exist, just the number of different titles. So, even if there’s only one copy of most books and 100 copies of one, they all count the same. 

Here, in the context of microbial communities, we are pondering: **“How many different ASVs are present?”** We do not weight them by abundance/evenness in any way. Rather, we focus on a binary measures of 0s and 1s that represent presence (1) and absence (0).

- Baker A’s starter has 10 unique ASVs: \( {}^0D_{\text{A}} = 10 \)    
- Baker B’s has 7 unique ASVs: \( {}^0D_{\text{B}} = 7 \)  
- Baker C’s has 15 unique ASVs: \( {}^0D_{\text{C}} = 15 \)  


### When \( q = 1 \): Balancing Rarity and Commonness  
> *Shannon Diversity* → \( {}^1D \)

Now we ask: **"How evenly are the ASVs distributed?"** Rare ASVs matter, but not as much as common ones. If we revisit the library analogy, now imagine you’re randomly picking books off the shelves to read. Some books have many more copies, so you’re more likely to pick those — but you still have a chance to encounter the rare ones. This measure reflects both richness and evenness — it tells you the effective number of “common” book titles based on their relative frequency. `q = 1` gives you a sense of how diverse your reading experience would be, considering popular vs. rare titles.

- Baker A’s 10 ASVs are 90% dominated by just one ASV: \( {}^1D_{\text{A}} \approx 2 \)  
- Baker B’s 7 ASVs are fairly even: \( {}^1D_{\text{B}} \approx 6.8 \)  
- Baker C’s 15 ASVs include 3 very dominant ones: \( {}^1D_{\text{C}} \approx 5 \)  

Even though Baker C and A have more ASVs in terms of richness in \( {}^0D \), **Baker B** has the highest \( {}^1D \) due to better balance between the abundances of its seven ASVs. 

### When \( q = 2 \): Focus on the Most Abundant ASVs  
> *Simpson-like Diversity* → \( {}^2D \)

Now we modify our question to wonder: **How many ASVs dominate each sample?** Here, rare ASVs hardly matter. For example, this is like walking into the library and only noticing the most abundant book titles, because they take up entire shelves. The rare titles are barely visible. Therefore, this measure emphasizes dominant species—those with many copies. `q = 2` tells us the effective number of dominant book titles, downplaying the presence of rare ones.

- Baker A’s starter is almost all one ASV of 10 total ASVs: \( {}^2D_{\text{A}} \approx 1.2 \)  
- Baker B’s 7 ASVs are balanced: \( {}^2D_{\text{B}} \approx 6.5 \)   
- Baker C’s has 3 dominant ASVs out of 15: \( {}^2D_{\text{C}} \approx 3.5 \)  

This tells us: **Baker B** has the most diverse community in terms of its *dominant* community members.


---

## Hill Number Summary Table

| Value of \( q \) | Emphasis                    | Hill Number \( {}^qD \) | Interpretation                              |
|------------------|-----------------------------|-------------------------|----------------------------------------------|
| \( q = 0 \)      | Every ASV counts equally     | \( {}^0D \)             | ASV richness — a simple count                |
| \( q = 1 \)      | Balance of common & rare     | \( {}^1D \)             | Shannon Diversity — balanced view            |
| \( q = 2 \)      | Dominant ASVs matter most    | \( {}^2D \)             | Simpson Diversity — focus on major players        |

---

So, hopefully you can see that using Hill numbers like \( {}^qD \) gives you a powerful, flexible way to describe microbial diversity — from just counting ASVs to understanding community structure. Next, we will calculate the biodiversity with the [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) package. iNEXT focuses on three measures of biodiversity via the Hill numbers. For each Hill NUmber [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) uses the observed abundance of species within a sample to compute diversity estimates and the associated 95% confidence intervals for the following two types of rarefaction and extrapolation (R/E):

1. **Sample-size-based (or size-based) R/E curves**: iNEXT computes diversity estimates for
rarefied and extrapolated samples up to an appropriate size (2x the total number of individuals/sequences in that sample). This type of sampling curve plots the diversity estimates with respect to sample size.
2. **Coverage‐based R/E curves:** iNEXT computes diversity estimates for rarefied and extrapolated samples based on a standardized level of sample completeness (as measured by sample coverage) up to an appropriate coverage value. This type of sampling curve plots the diversity estimates with respect to sample coverage. **NOTE: Coverage is inapplicable to DADA2 datasets, as they lack singletons** (this is also why we can’t use Amy Willis’s breakaway approach). 

*An important note:* We’ll be using the size_based output. Our lab has previously been suspicious of the Asymptotic Estimator. Therefore we will use the size_based approach here, with the caveat that we will very rigorously confirm that we’ve approached saturation for all of our samples! 


> Warning! iNEXT takes a while to run, especially if your samples are deeply sequenced or if you have many samples! It is wise to calculate iNEXT early and then come back to the rest of the home work later on.

```{r iNEXT-biodiversity}
# prepare the input data 
# Species in rows and the samples in columns 
iNEXT_input_df <- 
  preprocessed_physeq %>%
  otu_table() %>%
  data.frame()

# Inpect 
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

# Run iNEXT 
# Remember to set the seed!
# ASVs in ROWs, samples in COLUMNS 
iNEXT_data <- iNEXT(iNEXT_input_df, 
                    q = c(0, 1, 2),
                    datatype = "abundance")

# Save the object for later! 
# Make the directory first in shell! Otherwise there will be an error
save(iNEXT_data, file = "data/04_Biodiversity/iNEXT_data.RData")
```

Since iNEXT takes a while to run, we will load in the data that was caluculated before! 

```{r load-iNEXT-data}
load("data/04_Biodiversity_Lecture/iNEXT_data.RData")

# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)
```


# Rarefaction Curves 

## `ggiNEXT` Rarefaction Curves 

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()

# Check
head(color_df)

# Rename the column 
colnames(color_df)[1] <- "names"

# Check
head(color_df)
metadata_df

colnames(color_df)

colnames(metadata_df)

# Make a helper dataframe for plotting with colors 
group_colors <- c("China" = "red4")

iNEXT_color_df <- 
  color_df %>%
  mutate(
    names = gsub(pattern = "[.]", replace = "-", x = names),
    names = gsub(pattern = "^X", replace = "", x = names)
  ) %>%
  rename(SampleID = names) %>%
  left_join(metadata_df, by = "SampleID") %>%
  left_join(
    data.frame(
      group_colors = group_colors,
      geo_loc_name_country = names(group_colors)
    ),
    by = "geo_loc_name_country"
  )

head(iNEXT_color_df)
```

```{r plot-iNEXT-rarefaction, fig.width=6, fig.height=3.5}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$group_colors) + 
  scale_fill_manual(values = iNEXT_color_df$group_colors) + 
  theme_bw() + 
  theme(legend.position = "none") + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") 
```



