---
title: "Microbial Compositional Analysis"
author: "Jayden William Kgobe"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      fig.path = "../../figures/Lecture/06_Composition_Lecture/")
```

# Goals

In this file, we will perform compositional analysis of our scaled/noramlized/rarefied microbial dataset! 

1. Load in scaled/normalized phyloseq object. 
2. Calculate the relative abundances of taxonomic groups at various levels: 
    A. Phylum
    B. Genus
    C. ASV 
3. Plot them, and narrow in on specific taxnomic group of interest

## Inputs 

1. We will need the `scaled_physeq.RData`, which includes a rooted tree that we created in `analysis/05_Ordination/scaled_physeq.RData`. 

## Outputs 

1. Beautiful visualizations of microbial taxa and how they vary across parameters related to our study: station (categorical) and salinity (continuous).
2. Run some stats! 

# Compositional Data 

Microbial abundance data—like 16S rRNA gene or metagenomics data—are typically **compositional:** they represent relative abundances constrained to a constant total (*e.g.,* percent or proportions). This introduces spurious correlations and other issues if analyzed with traditional statistics. This is a very important limitation to microbial data! 
# Setup 

## Load Packages 
```{r load-packages}
# Load in the packages 
pacman::p_load(tidyverse, devtools, DT, phyloseq, patchwork, 
               install = FALSE)

# load colors
source("code/colors_and_shapes.R")
```

## 1. Load in Scaled Phyloseq object 

The following phyloseq object contains microbial community composition data in a standardized format. In this case, we’ve already normalized the reads (scaled to 1,942 per sample), which is essential for comparing relative abundances.


```{r load-data}
load("data/05_Ordination/scaled_physeq.RData")

# Look at the data 
scaled_physeq

# Intuition check - scaled at 1,942
min(sample_sums(scaled_physeq))
range(sample_sums(scaled_physeq))
```

# Taxonomic Analysis! 

In this analysis, we will drill down from phylum to ASV level analyses, which will enable increasingly detailed insights into microbial diversity and potential ecological roles. However, it is also important for us to remember that deeper levels also come with increased noise and data sparsity, especially for rare groups. 

## A. Phylum 

Taxonomic analysis often begins at broader levels (*e.g.,* Phylum) to visualize overarching trends before zooming in on finer-scale patterns. This step allows us to identify which microbial groups dominate across samples and which may respond to environmental gradients like salinity.

Now, let's calculate the relative abundance of the phyla across all the samples. **NOTE:** To do this, it is imperative that we have scaled the data to a constant sequencing depth. 

```{r calc-phylum-df}
# Create a phylum level dataframe
phylum_df <- 
  scaled_physeq %>%
  # Agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Phylum") %>%
  # Calculate the relative abundance! 
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # Create a dataframe from phyloseq object
  psmelt()
  # Filter out Phyla < 1 % 
  #dplyr::filter(Abundance > 0.01) %>%

## What are the phylum abundances? 
phylum_df %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = round(mean(Abundance), digits = 4)) %>%
  arrange(-mean_PercAbund) %>%
  datatable()

# Make a list of phyla the top phyla 
top10_phyla <- 
  phylum_df %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  head(n = 10) %>%
  pull(Phylum)
# show it 
top10_phyla
```

## Faceted Phylum Plot 

To help compare the phylum abundance between sample types, we can facet by phylum to better see how the changes occur across the groups, which is masked in the stacked bar plot. It's a little better than the stacked bar plot, however, we can do even better! 

```{r plot-phylum-station, fig.width=12, fig.height=6}
### Or combined together: 
phylum_df %>%
  dplyr::filter(Phylum %in% top10_phyla) %>%
  ggplot(aes(x = Group, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(Phylum~., scales = "free", nrow = 2) + 
  # add the stacked bar 
  geom_jitter() +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) + 
  # change the colors to be our selected colors 
  scale_fill_manual(values = phylum_colors) + 
  scale_color_manual(values = phylum_colors) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
```

## Stacked Bar Chart

```{r stacked-phylum-group, fig.width=8, fig.height=5}
# Create group-level phylum relative abundance dataframe
phylum_group_df <- 
  phylum_df %>%
  dplyr::filter(Phylum %in% top10_phyla) %>%
  group_by(Group, Phylum) %>%
  summarize(mean_Abundance = mean(Abundance), .groups = "drop")

# Stacked bar plot by Group
phylum_group_df %>%
  ggplot(aes(x = Group, y = mean_Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = phylum_colors) +
  labs(
    x = "Group",
    y = "Mean Relative Abundance",
    fill = "Phylum"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

```






