---
title: "Microbial Compositional Analysis"
author: "Jayden William Kgobe"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      fig.path = "../../figures/06_Composition/")
```

# Goals

In this file, we will perform compositional analysis of our scaled/noramlized/rarefied microbial dataset! 

1. Load in scaled/normalized phyloseq object. 
2. Calculate the relative abundances of taxonomic groups at various levels: 
    A. Phylum
    B. Genus
    C. ASV 
3. Plot them, and narrow in on specific taxnomic group of interest

## Inputs 

1. We will need the `scaled_physeq.RData`, which includes a rooted tree that we created in `analysis/05_Ordination/scaled_physeq.RData`. 

## Outputs 

1. Beautiful visualizations of microbial taxa and how they vary across parameters related to our study: station (categorical) and salinity (continuous).
2. Run some stats! 

# Compositional Data 

Microbial abundance data—like 16S rRNA gene or metagenomics data—are typically **compositional:** they represent relative abundances constrained to a constant total (*e.g.,* percent or proportions). This introduces spurious correlations and other issues if analyzed with traditional statistics. This is a very important limitation to microbial data! 
# Setup 

## Load Packages 
```{r load-packages}
# Load in the packages 
pacman::p_load(tidyverse, devtools, DT, phyloseq, patchwork, 
               install = FALSE)

# load colors
source("code/colors_and_shapes.R")
```

## 1. Load in Scaled Phyloseq object 

The following phyloseq object contains microbial community composition data in a standardized format. In this case, we’ve already normalized the reads (scaled to 1,942 per sample), which is essential for comparing relative abundances.


```{r load-data}
load("data/05_Ordination/scaled_physeq.RData")

# Look at the data 
scaled_physeq

# Intuition check - scaled at 1,942
min(sample_sums(scaled_physeq))
range(sample_sums(scaled_physeq))
```

# Taxonomic Analysis! 

In this analysis, we will drill down from phylum to ASV level analyses, which will enable increasingly detailed insights into microbial diversity and potential ecological roles. However, it is also important for us to remember that deeper levels also come with increased noise and data sparsity, especially for rare groups. 

## A. Phylum 

Taxonomic analysis often begins at broader levels (*e.g.,* Phylum) to visualize overarching trends before zooming in on finer-scale patterns. This step allows us to identify which microbial groups dominate across samples and which may respond to environmental gradients like salinity.

Now, let's calculate the relative abundance of the phyla across all the samples. **NOTE:** To do this, it is imperative that we have scaled the data to a constant sequencing depth. 

```{r calc-phylum-df}
# Create a phylum level dataframe
phylum_df <- 
  scaled_physeq %>%
  # Agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Phylum") %>%
  # Calculate the relative abundance! 
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # Create a dataframe from phyloseq object
  psmelt()
  # Filter out Phyla < 1 % 
  #dplyr::filter(Abundance > 0.01) %>%

## What are the phylum abundances? 
phylum_df %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = round(mean(Abundance), digits = 4)) %>%
  arrange(-mean_PercAbund) %>%
  datatable()

# Make a list of phyla the top phyla 
top10_phyla <- 
  phylum_df %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  head(n = 10) %>%
  pull(Phylum)
# show it 
top10_phyla
```

## Faceted Phylum Plot 

To help compare the phylum abundance between sample types, we can facet by phylum to better see how the changes occur across the groups, which is masked in the stacked bar plot. It's a little better than the stacked bar plot, however, we can do even better! 

```{r plot-phylum-station, fig.width=12, fig.height=6}
### Or combined together: 
phylum_df %>%
  dplyr::filter(Phylum %in% top10_phyla) %>%
  ggplot(aes(x = Group, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(Phylum~., scales = "free", nrow = 2) + 
  # add the stacked bar 
  geom_jitter() +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) + 
  # change the colors to be our selected colors 
  scale_fill_manual(values = phylum_colors) + 
  scale_color_manual(values = phylum_colors) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
```

## Stacked Bar Chart

```{r stacked-phylum-group, fig.width=8, fig.height=5}
# Create group-level phylum relative abundance dataframe
phylum_group_df <- 
  phylum_df %>%
  dplyr::filter(Phylum %in% top10_phyla) %>%
  group_by(Group, Phylum) %>%
  summarize(mean_Abundance = mean(Abundance), .groups = "drop")

# Stacked bar plot by Group
phylum_group_df %>%
  ggplot(aes(x = Group, y = mean_Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = phylum_colors) +
  labs(
    x = "Group",
    y = "Mean Relative Abundance",
    fill = "Phylum"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

```
## A2: Top 10 ASVs

Now, let's take a look at the ASVs. This is where the *real* ecology/biology happens! This is because the ASV-level plots will provide us with a more detailed view of **which specific taxa (ASVs) are driving the overall trends seen at the higher taxonomic levels (e.g. phylum)** in relation to salinity gradients and station differences.

Before we calculate the ASV-level abundances, let's take a second to think about *who* we'd like to consider. There's a lot of ASVs! In fact, there are `r ntaxa(scaled_physeq)`! It is difficult analyze all of them! Therefore, we will remove ASVs that have an overall count across the entire dataset of 211 or 1% of a scaled sample to 2,107 This is a little arbitrary and therefore, you may choose a different threshold for your dataset. The goal here is to **dramatically decreases the number of ASVs in the dataset** to help make the ASV-level data analysis much easier. 

Of course, if we had more time in class, we could run differential abundance and this would also help us to identify and statistically test which ASVs are important and also help us narrow on specific ASVs. 

The goal with this lesson is to do some data exploration and then also test ASVs that are relevant for our specific study. Let's go! 

```{r calc-ASVs, fig.width=12, fig.height=4}
# Calculate the Family relative abundance 
# Note: The read depth MUST be normalized in some way: scale_reads
ASV_df <- 
  scaled_physeq %>%
  # Prune out ASVs that have fewer than ~1%! 
  ## LOOK AT HOW MANY ARE REMOVED! We scaled to 2,107 reads! 
  prune_taxa(taxa_sums(.) >= 211, .) %>%
  # agglomerate at the phylum level 
  tax_glom(taxrank = "ASV") %>% 
  # Transform counts to relative abundance 
  transform_sample_counts(function (x) {x/sum(x)}) %>%
  # Melt to a long format 
  psmelt() 

# Make a list of phyla the top 10 ASVs in the Dataset
top10_ASVs <- 
  ASV_df %>%
  group_by(ASV) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  head(n = 10) %>%
  pull(ASV)

# show it 
top10_ASVs
```

```{r plot-Top10-ASVs-salGroup, fig.width=12, fig.height=9}
### Plot the Top 10 ASVs
ASV_df %>%
  dplyr::filter(ASV %in% top10_ASVs) %>%
  ggplot(aes(x = Group, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(ASV~Genus, scales = "free", nrow = 3) + 
  # add the stacked bar 
  geom_jitter() +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) + 
  # change the colors to be our selected colors 
  scale_fill_manual(values = phylum_colors) + 
  scale_color_manual(values = phylum_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        axis.title.x = element_blank())
```
# B. ASV-Level

## B1. Bacteroidota
```{r bacter-asvs, fig.width=10, fig.height=8}
# Calculate top couple of ASVs 
# Make a list of phyla the top phyla 
top_bacter_ASVs <- 
  ASV_df %>%
  dplyr::filter(Phylum == "Bacteroidota") %>%
  group_by(ASV) %>%
  summarize(mean_Abundance = mean(Abundance)) %>%
  dplyr::filter(mean_Abundance > 0.005) %>%
  pull(ASV)

# Bacteroidota
# Plot ASVs 
bacter_asv_Group <- 
  ASV_df %>%
  dplyr::filter(ASV %in% top_bacter_ASVs) %>%
  # build the plot 
  ggplot(aes(x = Group, y = Abundance, 
             fill = Group, color = Group)) + 
  facet_wrap(Genus~ASV, scales = "free_y", nrow = 2) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + # outliers not plotted here in boxplot 
  geom_jitter() + 
  theme_bw() + 
  labs(title = "Bacteroidota ASVs") + 
  scale_color_manual(values = group_colors) + 
  scale_fill_manual(values = group_colors) + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        legend.position = "none")


# Collect the Bacteroidota Plots
bacter_asv_Group  + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```



