---
title: "02_AssignASVs"
author: "Jayden William Kgobe"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/02_AssignASVs")
```

# Goals 

1. Infer an error model for in the filtered sequences, separately on forward and reverse reads.
2. Assign ASVs on both forward and reverse reads separately by applying the error model.
3. Merge forward and reverse ASVs into "contiguous ASVs".
4. Generate the first draft of ASV count table.
5. Quality Trimming of ASV lengths.
6. Remove chimeras. 
7. Assign Taxonomy with Silva Database. 
8. Write out relevant files: `asv_table`, `asvs_fasta`, `tax_table`, and `sample_data`.
9. Create a **S4** R data object, which in this case is known as a phyloseq, object from the `asv_table`, `tax_table`, and metadata. This will give us one data object to work with moving forward. :)

## Input 

1. Filtered fastq files generated from `01_QualityTrimming.Rmd`.
2. Sample Name vector.

## Output 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta` for building a phylogenetic tree at a later step.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 

# Set up the Environment 

## Set Seed & Threads
```{r set-seed}
# Set the seed for reproducibility
set.seed(40125)

# Let's make a parameter to set the number of threads 
n_threads = 30
```

## Timing of this script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load Packages 
```{r load-packages}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, devtools, dada2, phyloseq, 
               patchwork, DT, install = FALSE)
```

## Load Filtered Fastq Files 

```{r load-filtered-fastqs}
# Place filtered seq files into a variable 
filtered_fastqs_path <- "data/01_QualityTrimming/filtered_fastqs"

# Intuition check:
filtered_fastqs_path

# Create Forward vector 
filtered_forward_reads <- 
  list.files(filtered_fastqs_path, pattern = "1_filtered.fastq.gz",
             full.names = TRUE)
# Check 
filtered_forward_reads[1:5]

# Reverse vector 
filtered_reverse_reads <- 
    list.files(filtered_fastqs_path, pattern = "2_filtered.fastq.gz",
             full.names = TRUE)

# Check 
filtered_reverse_reads[1:5]
```
## Sample Names 
```{r sample-names}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forward_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)
```

# Error Modelling 

## How does it work? 

This is the step along the workflow where we try to estimate what is a *mistake* from technological sequencing error versus what is *true biological variation*. Therefore, this step is another critical step that we need to be mindful of! This is also what makes DADA2 unique!  

Specifically, we will infer error rates for all possible *transitions* within purines and pyrimidines (A<>G or C<>T) and *transversions* between all purine and pyrimidine combinations. The error model is learned by alternating estimation of the error rates and inference of sample composition until they converge. It starts with abundant sequences first and then goes to less abundant sequences. It will:  

1. Starts with the assumption that the error rates are the maximum (takes the most abundant sequence ("center") and assumes it's the only sequence not caused by errors).  
2. Compares the other sequences to the most abundant sequence. 
3. Uses at most 10^8^ nucleotides for the error estimation. (Though, sometimes we increase this parameter in the case of binned sequencing quality scores.)  
4. Uses parametric error estimation function of loess fit on the observed error rates. 

## Important Notes

**First, every individual sequencing run needs to have it's own and separate error model!** This is because every sequencing run is unique—even if the flow cell and DNA/samples are *exactly the same*. 

While you can combine multiple datasets together after running them through DADA2, if you'd like to limit bias and combine data sets from multiple sequencing runs, it is ideal to do the following:  

1. Use the exact same DNA extractions and library prep (*e.g.* same primer sets). 
2. Perform the exact same `filterAndTrim()`, with equivalent parameters between sequencing runs. This is because we need to have the exact same ASV length expected by the output of our filtering and trimming.
3. Assign ASVs on forward and reverse reads, merge them, and then once you have the ASV table from each sequencing run, you can merge them together.

But wait: *What contributes to sequencing error in different sequencing runs and why do we need to model errors separately per run with `learnErrors()` in dada2?* Remember the core principles of how Illumina sequencing works! Some things that contribute to this are:  

a. Each sequencing run has a **unique timing for when clusters go out of sync** (*e.g.,* the drop in quality at end of reads, which is typical of Illumina sequencing. This decrease in read quality is especially true of older Illumina sequencing data and is improving with more recent technologies. 
b. The **cluster density is impossible to replicate**. Therefore, the cluster density (and therefore sequence quality) will always be different between sequencing runs (even if it's the same person/samples/sequencing facility!). This is getting easier to replicate with the newer flow cells (*e.g.,* patterned flow cells) but is not yet possible. 
c. **PhiX spike-in will also vary between runs**, even if we try to make it the same! Therefore, the amount of heterogeneity on the flow cell will also be different, which further impacting the sequencing quality.  
d. **Different locations on the flow cell can be impacted differently between runs.** Perhaps an air bubble can get in. Or, as mentioned above, the cluster density is higher and/or lower on a different run/flow cell. In the old, non-patterned flow cells, sometimes clusters can form nearby each other. However, this has been solved on the new, patterned flow cells where clusters have a specific distance between each other. 

**Second, this step uses our quality scores, which is rapidly changing with new sequencing technologies!** Therefore, we need to be mindful of how we run our error models. The traditional error models were developed based on the 40 Phred scores that were generated by early Illumina Sequencing technologies. However, with the new binned quality scores, we need to incorporate new ways of error modeling and this is constanstly changing right now, which you can see in this [very active discussion on the DADA2 GitHub page](https://github.com/benjjneb/dada2/issues/1307). 

Now, let's consider the two types of Illumina data output and how it might inform our error modeling. 

## Learn the Errors
### MiSeq Runs: 40 Phred Scores

Here, in the data we have sequenced here, we have Illumina MiSeq data with the traditional 40 Phred scores. Therefore, we can use the traditional `learnErrors()` command, which we will do below. 

```{r learn-errors-MiSeq, fig.width=12, fig.height=8}
# Forward Reads 
error_forward_reads <- 
  learnErrors(filtered_forward_reads, multithread = n_threads)

# Forward Error Plot 
forward_error_plot <- 
  plotErrors(error_forward_reads, nominalQ = TRUE) + 
  labs(title = "Forward Reads: Error Model")

# Reverse Reads 
error_reverse_reads <- 
  learnErrors(filtered_reverse_reads, multithread = n_threads)

# Reverse Error Plot 
reverse_error_plot <- 
  plotErrors(error_reverse_reads, nominalQ = TRUE) + 
  labs(title = "Reverse Reads: Error Model")

# Look at the plots together 
forward_error_plot + reverse_error_plot
```

The above plot represents the error rates for each possible transition (A→C, A→G, …) in the forward reads (on the left) and the reverse reads (on the right).

Details of the plots above: 
- **Points**: The observed error rates for each consensus quality score.  
- **Black line**: Estimated error rates after convergence of the machine-learning algorithm.  
- **Red line:** The error rates expected under the nominal definition of the Q-score.  

Similar to what is mentioned in the [DADA2 tutorial](https://benjjneb.github.io/dada2/tutorial_1_8.html): the estimated error rates (black line) are a "reasonably good" fit to the observed rates (points), and the error rates drop with increased quality as expected.  We can now infer ASVs! 

# Infer ASVs

**An important note:** What makes DADA2 so nice is that it infers ASVs separately on the forward and the reverse reads and then merges them afterwards. This is quite a different approach from how OTUs are identified in Mothur and also from UCHIME, oligotyping, and other OTU, MED, and ASV approaches.

```{r infer-ASVs}
# Infer ASVs on the forward sequences
dada_forward <- 
  dada(filtered_forward_reads, 
       err = error_forward_reads,
       multithread = n_threads) 

# Take a look at the data
# What type of data structure is it? 
typeof(dada_forward) # It's a list 
length(dada_forward) # How big is it? One per sample!

# What doees it look like for each sample?  
dada_forward$`SRR12873151_1_filtered.fastq.gz`

# Reverse ASVs
dada_reverse <- 
  dada(filtered_reverse_reads,
       err = error_reverse_reads ,
       multithread = n_threads)

# Check data 
dada_reverse[30]
```


# Merge Forward and Reverse ASVs 

Now, that we have identified our ASVs separately on both the forward and the reverse reads, let's **merge** them together into contiguous (*aka* "contigs") ASVs.

```{r merge-ASVs}
merged_ASVs <- 
  mergePairs(dada_forward, filtered_forward_reads,
             dada_reverse, filtered_reverse_reads,
             verbose = TRUE)

# Evaluate the data output 
typeof(merged_ASVs) # A list
length(merged_ASVs) # Length of the number of samples!
head(names(merged_ASVs)) # Here, we can access our current sample names

# Inspect further for each sample
#head(merged_ASVs, n = 2) # A dataframe for each sample
# We have a dataframe in each part of our list! What are in the columns? 
glimpse(merged_ASVs$`SRR12873151_1_filtered.fastq.gz`)
```

# Create Raw ASV Count Table 

```{r raw-ASV-count-table}
# Raw ASV
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Intuition Check: Type and dimensions of the data
dim(raw_ASV_table)
typeof(raw_ASV_table)
class(raw_ASV_table)

# write out raw_asv_table 
write.table(raw_ASV_table, file = "data/02_AssignASVs/raw_ASV_counts.tsv",
            sep = "\t", quote = FALSE, col.names = NA)
```

# Assess ASV Quality 

## What are the current ASV lengths? 

First, let's inspect the distribution of the ASV sequence lengths across all of the ASVs that we have inferred. First, we need to consider: **What is the expected length of the ASVs?** 

1. *What is the sequencing strategy?*
    - paired-end 250bp (2x300bp) Illumina MiSeq Sequencing was used.
    - Specifically, the V3-V4 hypervariable region of the 16S rRNA gene was targeted for sequencing with the 338F and 806R primers. 
2. *What's the total length of our starting amplicons?* 
    - Our primers are named after the location they bind to on the 16S gene. So: 
    - `806 - 338 = 468` 
    - The total length of the starting amplicon is 468 base pairs, with primers.
3. *What is the ASV length without primers?* 
    - The 338F (`ACTCCTACGGGAGGCAGCAG`) primer is 20 base pairs. 
    - The 806R (`GGACTACHVGGGTWTCTAAT`) primer is 20 base pairs. 
    - `468 - 20 - 20 = 428` 
    - The expected ASV length without primers is **428 base pairs**. 
    - **IMPORTANT:** Remember, we did not sequence the primers in our current example, because the [Kozich et al., 2013](https://journals.asm.org/doi/10.1128/aem.01043-13) library prep was used. However, if the regular Illumina 2-step library preparation approach was used, the primers will be sequenced!
4. *What is the length of the trimmed ASVs after `filterAndTrim()`?* 
    - In our `01_QualityTrimming.Rmd`, we used `trimLeft = 20` to remove the bases at the beginning of the forward read and `truncLen = c(280,250)`, which removed 20 bases from the forward read length and removed 50 bases at the end of the reverse read. Therefore: 
    - `428 - 20 = 408`
5. *What is the overlap between our forward and reverse reads?*
    - We performed 2x300 paired-end Illumina MiSeq Sequencing. However, from our multiQC report, we can see that our read length is actually 251 bp. 
    - If we have a total read length of 251 base pairs and **we did not sequence our primers** (because we used the Kozich et al., 2013 AEM library preparation), then our overlap should be ~100%. 
    
<span style="color: red;">INTERPRETATION #1: What is the hypothesized ASV length for your project data set? Please insert a description here, similar to the statements above, of the expected ASV length. Please provide a clear explanation using the following, which an example is provided above: (1) sequencing strategy, (2) total length of starting amplicons, (3) the expected ASV length without the primers, (4) length expected after the filterAndTrim() step, and (5) the expected percent overlap.</span>

### ASV Length Stats 

```{r assess-ASV-length}
# Calculate summary stats
# Longest ASV?
maxLength_ASV <- max(nchar(getSequences(raw_ASV_table)))

# Shortest ASV?
minLength_ASV <- min(nchar(getSequences(raw_ASV_table))) 

# Mean ASV length?
meanLength_ASV <- mean(nchar(getSequences(raw_ASV_table))) 

# Median ASV length?
medianLength_ASV <- median(nchar(getSequences(raw_ASV_table))) 

# Create a table to Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table)))

meanLength_ASV
```

In the table above, we can see that we actually have quite the ASV sequence length distribution! There are some longer ASVs that are clearly spurious... we need to be **skeptical**! We see that the most abundant ASV length is 245, which has order an of magnitude more ASVs than compared to any other length. Since the difference in ASV abundance at 245 is so much higher than any other ASV length, we will remove ASVs that are not 245 bp. 

Here's some of the stats that we've calculated:

- Max ASV length: `r maxLength_ASV` base pairs
- Min ASV length: `r minLength_ASV` base pairs
- Mean ASV length: `r meanLength_ASV` base pairs
- Median ASV length: `r medianLength_ASV` base pairs

However, the median length actually is about what we are expecting, which is great. *Let's take a look at what this looks like graphically!*

### ASV Length Plot 

Now, we will plot these ASV lengths to give us a better idea: 

```{r ASV-lengths-plot}
# Inspect the distribution of sequence lengths of all ASVs in data set 
# AFTER TRIM
plot_ASVLength_raw <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Raw ASV Length (bps)")

# Show the plot
plot_ASVLength_raw
```

As we saw in the table, the length of the ASVs is mostly 245 base pairs. 

<span style="color: red;">INTERPRETATION #2: Does the ASV length data (table and plot above) match your hypotheses from INTERPRETATION #1? Why or why not? Please insert a description of the interpretation you draw regarding the lengths of your ASVs that you've inferred from the ASV length table and it's graphical visualization above. </span>

***

<span style="color: red;">INTERPRETATION #3: Taking into account INTERPRETATIONS #1 and #2, what is your suggested trimming procedure for your ASV lengths? Will you allow some variation or trim to an exact length, similar to our in class example here?</span>

## Trim ASV lengths 

Now that we are pretty convinced that we have some spurious ASVs, we will trim ASVs that are higher or lower than 245 base pairs in this data set. 

**NOTE:** This is another moment in our workflow where we are sub-setting only the "good parts" of our data to ensure the data quality we have at the end is the best of our ability based on the sequencing run. 

*Let's trim the ASVs to only be the right size, which is 245!* 

```{r trim-ASVs}
# Subset only ASVs that are 245 bps long 
raw_ASV_table_trimmed <- 
  raw_ASV_table[,nchar(colnames(raw_ASV_table)) ==428]

# Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table_trimmed)))

# What proportion of total ASV sequences are left in the data? 
percRetained_Trimmed <- sum(raw_ASV_table_trimmed)/sum(raw_ASV_table)
percRetained_Trimmed # Show it 

# Inspect the distribution of sequence lengths of all ASVs in dataset 
# AFTER TRIM
plot_ASVLength_trimmed <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Trimmed ASV Length (bps)")

# Show the plot 
plot_ASVLength_trimmed
```

